---
title: "ClimateR package"
output: html_document
editor_options: 
  chunk_output_type: console
---

From: https://mikejohnson51.github.io/climateR-intro

```{r}
library(rgdal)
library(sf)
library(tidyverse)
library(lubridate)
library(ggthemes)

# remotes::install_github("mikejohnson51/AOI") # suggested!
# remotes::install_github("mikejohnson51/climateR")
library(climateR)
library(AOI)
```

Get grid MET data. http://www.climatologylab.org/gridmet.html

What variables are available?
```{r}
param_meta$gridmet
```

Get a single variable, one day, over one state.
```{r}
FL.sf <- aoi_get(state = "FL")
plot(FL.sf$geometry)

rain <- getGridMET(FL.sf, 
                   param = "prcp", 
                   startDate = "2020-03-10")
class(rain$prcp)

plot(rain$prcp)
plot(FL.sf$geometry, add = TRUE)
```

Get the 1000 hr dead fuel moisture.
```{r}
data <- getGridMET(aoi_get(state = "FL"), 
                   param = "fmoist_1000", 
                   startDate = "2020-03-10")

plot(data$fmoist_1000)
plot(FL.sf$geometry, add = TRUE)
```

Point based data
```{r}
AOI <- AOI::geocode('Tallahassee, Florida', pt = TRUE)
ts  <- getGridMET(AOI, param = 'tmax', startDate = "2020-01-01", endDate = "2020-11-30")

ggplot(data = ts) + 
  aes(x = date, y = ( tmax - 273.16 ) ) + 
  geom_line() +
  theme_linedraw() + 
  labs(title = "Daily High Temperature: Tallahassee, FL 2020", x = "Date", y = "Temperature (C)")
```

Get for ANF boundaries.
```{r}
# download.file(url = "https://www.dropbox.com/s/wo52hwy6ol83nt8/NorthFloridaPublicLandHexagons10km.zip?dl=1",
#               destfile = "data/TempPubLand.zip")
# unzip("data/TempPubLand.zip", exdir = "data")

LandHex.sf <- read_sf(dsn = "data/NorthFloridaPublicLandHexagons10km") %>%
  dplyr::filter(NAME == "APALACHICOLA")
```

Get grid met data.
```{r}
data <- getGridMET(LandHex.sf, 
                   param = "fmoist_1000", 
                   startDate = "2020-03-10")

plot(data$fmoist_1000)
```

Convert to a {stars} object
```{r}
library(stars)
data.st <- st_as_stars(data$fmoist_1000)

data.st
```

Map
```{r}
library(tmap)

tm_shape(data.st) +
  tm_raster() +
tm_shape(LandHex.sf) +
  tm_borders()
```

One variable, 31 days.
```{r}
data <- getGridMET(LandHex.sf, 
                   param = "fmoist_1000", 
                   startDate = "2020-03-01",
                   endDate = "2020-03-31")
data.st <- st_as_stars(data$fmoist_1000)

names(data.st) <- "fmoist_1000"

days <- seq(as_date("2020-03-01"), 
            as_date("2020-03-31"), 
            by = "day")

( data.st <- st_set_dimensions(data.st, 
                               which = 3, 
                               values = days, names = "date") )

LandHexLL.sf <- LandHex.sf %>%
  st_transform(crs = st_crs(data.st))

X <- st_crop(data.st, 
             LandHexLL.sf, 
             crop = TRUE)

ggplot() +  
  geom_stars(data = X) + 
  facet_wrap("date") +
  scale_fill_viridis_c() +
  coord_sf() +
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm"))
```

Two variables, 31 days.
```{r}
data <- getGridMET(LandHex.sf, 
                   param = c("prcp", "tmax"),
                   startDate = "2019-03-01",
                   endDate = "2019-03-31")

data.st <- st_as_stars(data$prcp)
names(data.st) <- "prcp"

days <- seq(as_date("2020-03-01"), 
            as_date("2020-03-31"), 
            by = "day")

data.st <- st_set_dimensions(data.st, 
                             which = 3, 
                             values = days, names = "date")

( data.st <- data.st %>%
  mutate(tmax = ( as.vector(data$tmax) - 273.16 ) * 9/5 + 32,
         prcp = if_else(prcp < 0, 0, prcp * .03937008)) )

ggplot() +  
  geom_stars(data = data.st[1]) + 
  facet_wrap("date") +
  scale_fill_viridis_c() +
  coord_sf() +
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm"))
```

Compute KDBI for each grid point

Start with computing net rainfall.
```{r}
Q <- 269  
R <- 59.23 # average annual rainfall for TLH

NetRst <- array(data = rep(0, times = 30 * 17 * 31), dim = c(30, 17, 31))
Qlst <- array(data = rep(0, times = 30 * 17 * 31), dim = c(30, 17, 31))

for(x in 1:30){
  for(y in 1:17){

Rainfall24 <- data.st$prcp[x, y, ]
MaxTemp <- data.st$tmax[x, y, ]
for(i in 1:31){
  if(is.na(Rainfall24[i])) Rainfall24[i] = 0
  if(is.na(MaxTemp[i])) MaxTemp[i] = mean(data.st$tmax[, , ], na.rm = TRUE)
}

PR <- lag(Rainfall24)
PR[1] <- 0

CumR <- 0
NetR <- numeric()

Ql <- numeric()
DeltaQl <- numeric()

for(i in 1:length(Rainfall24)) {
  R24 <- Rainfall24[i]
  if (R24 == 0 | is.na(R24)) {
    NetR[i] <- 0
    CumR <- 0
  } 
  else if(R24 > 0 & R24 <= .2) {
      CumR <- CumR + R24
      if (PR[i] > .2 | CumR > .2) NetR[i] <- R24
      else if (CumR > .2) NetR[i] <- CumR - .2
      else NetR[i] <- 0
    }
  else if (R24 > .2) {
      if (CumR <= .2) {
      NetR[i] <- CumR + R24 - .2
      CumR <- CumR + R24
      }
      else {
      NetR[i] <- R24
      CumR <- CumR + R24
      }
  }
  DeltaQ <- (800 - Q) * (.968 * exp(.0486 * MaxTemp[i]) - 8.3) /(1 + 10.88 * exp(-.0441 * R)) * .001 
  Q <- ifelse(NetR[i] == 0,  Q + DeltaQ,  (Q + DeltaQ) - NetR[i] * 100)
  Q <- ifelse(Q < 0, 0, Q) 
  Ql <- c(Ql, Q)
}

NetRst[x, y, ] <- NetR
Qlst[x, y, ] <- Ql
   }
}
```

```{r}
data.st$NetR <- NetRst
data.st$Ql <- Qlst

ggplot() +  
  geom_stars(data = data.st[4]) + 
  facet_wrap("date") +
  scale_fill_viridis_c() +
  coord_sf() +
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm"))

LandHexLL.sf <- LandHex.sf %>%
  st_transform(crs = st_crs(data.st))

X <- st_crop(data.st, 
             LandHexLL.sf, 
             crop = TRUE)

ggplot() +  
  geom_stars(data = X[4]) + 
  facet_wrap("date") +
  scale_fill_viridis_c() +
  coord_sf() +
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm"))
```

Add the 1000 hr (and 100 hr) dead fuel moisture and compare with the Ql (KDBI)
```{r}
data <- getGridMET(LandHex.sf, 
                   param = "fmoist_100", 
                   startDate = "2020-03-01",
                   endDate = "2020-03-31")
( data.st <- data.st %>%
  mutate(fmoist_100 = as.vector(data$fmoist_100)) )

X <- st_crop(data.st, 
             LandHexLL.sf, 
             crop = TRUE)

cor.test(X$Ql, X$fmoist_100)
```

The correlation between KDBI and dead fuel moisture index is stronger with 1000 hr than with 100 hr