# Space-Time

https://rpubs.com/jelsner/tornadoRisk_longTermView
https://rpubs.com/jelsner/tornadoRisk_shortTermView

Focus on the Southeast U.S. Decrease the grid size.
Update with stars package? No because stars doesnt work with rasters.

Constructing space-time data frames. Vector and raster examples.

```{r}
sfdfL <- st_read(dsn = "1950-2018-torn-aspath") %>%
  filter(yr == 2014, st == "FL")
```

Next we project the geographic coordinate reference system to a specific Lambert conic conformal projection. The spatial units are set to meters.
```{r}
sfdfL <- st_transform(sfdfL, 
                      crs = "+proj=lcc +lat_1=60 +lat_2=30 +lon_0=-90 +units=m")
```

Next we add a buffer to the geometries to represent the tornado path ('footprint'). The width of the buffer is 1/2 the width given in the attribute table in the column `wid`. First we create new variables giving the width (and length) in units of meters.
```{r}
sfdfL <- sfdfL %>%
  mutate(Width = wid * .9144,
         Length = len * 1609.34)
sfdfB <- st_buffer(sfdfL, 
                   dist = sfdfL$Width/2,
                   endCapStyle = 'ROUND')
```

Next set the raster domain slightly larger than the bounding box and assign a resolution of one degree in longitude and one degree in latitude. Check the extent of the raster with the `extent()` function.
```{r}
r <- raster(xmn = -125, xmx = -67, ymn = 24, ymx = 50)
res(r) <- 1

extent(r)
```

Use the `rasterize()` function to count the number of times each raster cell contains a tornado genesis location. The first argument is the spatial data frame (dough; here a simple feature data frame) and the second is the raster without values (cooking cutter). The argument `field =` specifies a column name in the spatial data frame (here just an identifier) and the argument `fun =` specifies what to do (here simply count the unique instances of the field in each cell).
```{r}
Tor_r <- rasterize(Alltors, r, 
                   field = "om", 
                   fun = "count",
                   background = 0)

class(Tor_r)
dim(Tor_r)
```

The result is a raster layer. The number of tornadoes occurring in each cell are the values.
```{r}
values(Tor_r)[1:200]
```

Raster cells without tornadoes are given a value of `NA` the integers are the number of times the cell contained a tornado.

We make a quick map with the `plot()` method.
```{r}
plot(Tor_r)
```

It looks right. Some cells across the Plains and the South have quite a few tornadoes others not as many.

A spatial statistic indicating how similar values in neighboring cells tend to be is Moran's I. It is implemented on rasters with the `Moran()` function. 

First we convert the `NA` values to zero since they do not really represent missing information but rather no tornadoes. We do this by including a `background =` argument specifying a value of 0.
```{r}
Tor_r <- rasterize(Alltors, r, 
                   field = "om", 
                   fun = "count",
                   background = 0)
```

```{r}
Moran(Tor_r)
```