# Latent Field Modeling using INLA

In this chapter I show you how to fit a geostatistical model to predict malaria prevalence in The Gambia using the stochastic partial differential equation (SPDE) approach and the R-INLA package [@RueEtAl2014]. The chapter is adopted from Chapter 9 of @Moraga2020. https://www.paulamoraga.com/book-geospatial/sec-geostatisticaldataexamplespatial.html

We use data of malaria prevalence in children obtained at 65 villages in The Gambia which are contained in the {geoR} package [@RibeiroEtAl2020], and high-resolution environmental covariates downloaded with the {raster} package [@Hijmans2020]. We show how to create a triangulated mesh that covers The Gambia, the projection matrix and the data stacks to fit the model. 

Then we show how to manipulate the results to obtain the malaria prevalence predictions, and 95% credible intervals denoting uncertainty. We also show how to compute exceedance probabilities of prevalence being greater than a given threshold value of interest for policy making. 

Results are presented with interactive maps created with functions from the {tmap} package [@Tennekes2018].

### Prepare the data

First, we load the {geoR} package and attach the data `gambia` which contains information about malaria prevalence in children obtained at 65 villages in The Gambia.
```{r}
library(geoR)
data(gambia)
```

Next we inspect the data and see it is a data frame with 2035 observations and the following 8 variables:

`x`: x coordinate of the village (UTM),
`y`: y coordinate of the village (UTM),
`pos`: presence (1) or absence (0) of malaria in a blood sample taken from the child,
`age`: age of the child in days,
`netuse`: indicator variable denoting whether the child regularly sleeps under a bed net,
`treated`: indicator variable denoting whether the bed net is treated,
`green`: satellite-derived measure of the greenness of vegetation in the vicinity of the village,
`phc`: indicator variable denoting the presence or absence of a health center in the village.

```{r}
head(gambia)
```

Data in gambia are given at an individual level. Here, we do the analysis at the village level by aggregating the malaria tests by village. We create a data frame called d with columns containing, for each village, the longitude and latitude, the number of malaria tests, the number of positive tests, the prevalence, and the altitude.

#### Malaria prevalence

We can see that `gambia` has 2035 rows and the matrix of the unique coordinates has 65 rows. This indicates that 2035 malaria tests were conducted at 65 locations.
```{r}
dim(gambia)
dim(unique(gambia[, c("x", "y")]))
```

We create a data frame called `df` containing, for each village, the coordinates (`x`, `y`), the total number of tests performed (`total`), the number of positive tests (`positive`), and the malaria prevalence (`prev`). 

In data `gambia`, column `pos` indicates the tests results. Positive tests have `pos` equal to 1 and negative tests have `pos` equal to 0. Therefore, we can calculate the number of positive tests in each village by adding up the elements in `gambia$pos`. Then we calculate the prevalence in each village by calculating the proportion of positive tests (number of positive results divided by the total number of tests in each village). 

We create the data frame `df` using the {dplyr} package as follows:
```{r}
library(dplyr)

df <- gambia %>%
  group_by(x, y) %>%
  summarize(
    total = n(),
    positive = sum(pos),
    prev = positive / total
    )
head(df)
```

Next we create a simple feature data frame by making the `x` and `y` columns the spatial coordinates. We make a map of prevalence using functions from the {tmap} package.
```{r}
library(sf)

sfdf <- st_as_sf(x = df, 
                 coords = c("x", "y"),
                 crs = 32628)

library(tmap)
tmap_mode(mode = "view")

tm_shape(sfdf) +
  tm_dots(col = "prev", size = .1)
```

#### Environmental factors

We model malaria prevalence using a covariate that indicates the altitude in The Gambia. This covariate can be obtained with the `getData()` function of the {raster} package. This package can be used to obtain geographic data from anywhere in the world. 

In order to get the altitude values in The Gambia, we need to call `getData()` with the three following arguments:

* name of the data equal to "alt",
* country equal to the 3 letters of the International Organization for Standardization (ISO) code of The Gambia (GMB),
* mask equal to TRUE so the neighboring countries are set to NA.

```{r}
library(raster)

r <- getData(name = "alt", 
             country = "GMB", 
             mask = TRUE)
```

We make a map with the altitude raster.
```{r}
library(tmap)

tm_shape(r) +
  tm_raster() +
tm_shape(sfdf) +
  tm_dots(col = "prev", size = .1)
```

We add the the altitude values to the data frame `df` to be able to use it as a covariate in the model. Since the raster is longitude/latitude we convert the village locations to longitude/latitude adding them to the data frame.
```{r}
sfdfT <- sfdf %>%
  st_transform(crs = 4326)

df[, c("long", "lat")] <- st_coordinates(sfdfT)
```

Now we get the altitude values at the village locations using the `extract()` function from {raster}. The first argument of this function is the altitude raster (`r`). The second argument is a two-column matrix with the coordinates where we want to know the values, that is, the coordinates of the villages given by `d[, c("long", "lat")]`. We assign the altitude vector to the column `altitude` of the data frame `df`.
```{r}
df$altitude <- raster::extract(r, df[, c("long", "lat")])

head(df)
```

### Specify the model

Here we specify the model (mathematically and heuristically) to predict the prevalence of malaria in The Gambia using the stochastic partial differential equation (SPDE) approach and functions from the {INLA} package.

Mathematically, we assume that conditional on the true prevalence $P({\bf x}_i)$ at location ${\bf x}_i$, i = 1, $\ldots, n$, the number of positive results $Y_i$ out of $N_i$ people sampled follows a binomial distribution
$$
Y_i|P({\bf x}_i) \sim \hbox{Binomial}(N_i, P({\bf x}_i)) \\
\hbox{logit}[P({\bf x}_i)] = \beta_0 + \beta_1 \hbox{altitude} + S({\bf x}_i)
$$

Here $\beta_0$ is the intercept, $\beta_1$ is the coefficient on altitude and $S({\bf x}_i))$ is a spatial random effect (spatial autocorrelation) that follows a zero-mean Gaussian process with a Mat'ern covariance function
$$
\hbox{Cov}(S({\bf x}_i), S({\bf x}_j)) = \frac{\sigma^2}{2^{\nu-1}\Gamma(\nu)}(\kappa ||{\bf x}_i - {\bf x}_j||)^\nu K_\nu (\kappa ||{\bf x}_i - {\bf x}_j||)
$$
where $K_\nu$($\cdot$) is the modified Bessel function of the second kind with order $\nu$ > 0. $\nu$ is the smoothness parameters, $\sigma^2$ is the variance, and $\kappa$ > 0 is related to the practical range $\rho = \sqrt{8\nu}/\kappa$ which is the distance at which the spatial correlation is close to .1.

#### Constructing a mesh

First we extract the locations of the malaria cases as longitude and latitude coordinates.
```{r}
LonLat <- sfdf %>%
  st_transform(crs = 4326) %>%
  st_coordinates() 
head(LonLat)
```

We need to build a triangulated mesh using the longitude and latitude coordinates that covers The Gambia over which to make the random field discretization. We do this using Delauney triangulation. We use the `inla.mesh.2d()` function passing the following parameters:

* `loc`: location coordinates that are used as initial mesh vertices,
* `max.edge`: values denoting the maximum allowed triangle edge lengths in the region and in the extension,
* `cutoff`: minimum allowed distance between points.

Here, we call `inla.mesh.2d()` setting `loc` equal to the matrix with the coordinates `coo`. We set `max.edge = c(.1, 5)` to use small triangles within the region, and larger triangles in the extension. We also set `cutoff = .01` to avoid building many small triangles where we have some very close points.
```{r}
library(INLA)

coo <- cbind(LonLat[, 1], LonLat[, 2])
mesh <- inla.mesh.2d(loc = coo, 
                     max.edge = c(.1, 5),
                     cutoff = .01)
```

The number of the mesh vertices is given by `mesh$n` and we can plot the mesh with `plot(mesh)`.
```{r}
mesh$n

plot(mesh)
points(coo, col = "red")
```

#### Defining and indexing the spatial random effects term on the mesh 

Then, we use the `inla.spde2.matern()` function to define the spatial random effect on the mesh.
```{r}
sre <- inla.spde2.matern(mesh = mesh, 
                         alpha = 2, 
                         constr = TRUE)
```
Here, we set `constr = TRUE` to impose an integrate-to-zero constraint and `alpha is a parameter related to the smoothness parameter of the process, namely, $\alpha = \nu + d/2$. 

In this example, we set the smoothness parameter $\nu$ is set to 1 and in the spatial case $d$ = 2 so $alpha$ =  1 + 2/2 = 2.

Next we need to index each of the mesh nodes (vertices). We do this with the `inla.spde.make.index()` function where we specify the name of the effect and the number of nodes in the spatial random effect term (`sre$n.spde`).
```{r}
indexs <- inla.spde.make.index(name = "s",
                               n.spde = sre$n.spde)
lengths(indexs)
```

This creates a list with vector `s` equal to `1:sre$n.spde`, and vectors `s.group` and `s.repl` that have all elements equal to 1s and length given by the number of mesh nodes.

#### Projecting the observation locations to the mesh

Next we need to build a matrix A that projects the locations of the observations to the mesh nodes. The projection matrix is built with the `inla.spde.make.A()` function as follows.
```{r}
A <- inla.spde.make.A(mesh = mesh,
                      loc = coo)
```

#### Defining a prediction grid amd projecting the grid locations to the mesh

Next we need to define where we want the predictions to be made. The raster of elevations provides a set of locations and we can extract the raster cells as points with the `rasterToPoints()` function from the {raster} package. Since there are many cells we first reduce this number by a factor of five with the `aggregate()` function taking the mean elevation over five cells in each direction.
```{r}
ra <- raster::aggregate(r,
                        fact = 5,
                        fun = mean) 
dp <- rasterToPoints(ra)

dim(dp)
head(dp)
```

Next we extract just the spatial coordinates of the prediction locations and project these locations to the mesh.
```{r}
coop <- dp[, c("x", "y")]

Ap <- inla.spde.make.A(mesh = mesh,
                       loc = coop)
```

#### Organize the model inputs

We use the `inla.stack()` function to organize data, effects, and projection matrices with the following arguments:

* `tag`: string to identify the data,
* `data`: list of data vectors,
* `A`: list of projection matrices,
* `effects`: list with fixed and random effects.

We construct a stack called `stk.e` with data for estimation and we tag it with the string "est". 

The fixed effects are the intercept (b0) and a covariate (altitude). The random effect is the spatial Gaussian random field (s). Therefore, the `effects` argument gets a list with a data.frame with the fixed effects, and s containing the indices of the random effect (`indexs)`. 

The argument `A` is set to a list where the second element is `A`, the projection matrix for the random effects, and the first element is 1 to indicate the fixed effects are mapped one-to-one to the response. 

In the `data` argument we specify the response vector and the number of trials. 
```{r}
stk.e <- inla.stack(tag = "est",
                    data = list(y = df$positive, numtrials = df$total),
                    A = list(1, A),
                    effects = list(data.frame(b0 = 1, altitude = df$altitude), s = indexs))
```

We also construct a stack for prediction that called `stk.p`. This stack has tag equal to "pred", the response vector is set to NA, and the data is specified at the prediction locations. Note the `altitude` variable is in column 3 of the array `dp` from above. Finally, we put `stk.e` and `stk.p` together in a `stk.full`.
```{r}
stk.p <- inla.stack(tag = "pred",
                    data = list(y = NA, numtrials = NA),
                    A = list(1, Ap),
                    effects = list(data.frame(b0 = 1, altitude = dp[, 3]), s = indexs))

stk.full <- inla.stack(stk.e, stk.p)
```

### Fit the model

We specify the model formula by including the response on the left-hand side, and the fixed and random effects on the right-hand side. We remove the intercept (adding 0) and add it as a covariate term (adding `b0`) so that all the covariate terms can be captured in the projection matrix.
```{r}
formula <- y ~ 0 + b0 + altitude + f(s, model = sre)
```

Finally we fit the model by calling the `inla()` function and using the default priors. We specify the formula, family, data, and options. With the `control.predictor` argument we set `compute = TRUE` to compute the posteriors of the predictions. We set `link = 1` to compute the fitted values (`res$summary.fitted.values` and `res$marginals.fitted.values`) with the same link function as the family specified in the model.
```{r}
fit <- inla(formula,
            family = "binomial", 
            Ntrials = numtrials,
            control.family = list(link = "logit"),
            data = inla.stack.data(stk.full),
            control.predictor = list(compute = TRUE, link = 1, A = inla.stack.A(stk.full)))
```

### Map the results

The mean prevalence and lower and upper limits of 95% credible intervals are in the data frame `fit$summary.fitted.values`. The rows of the data frame correspond to the prediction locations and can be obtained by selecting the indices of the stack `stk.full` that are tagged with `tag = "pred"`.
```{r}
index <- inla.stack.index(stack = stk.full, tag = "pred")$data
```

We create vectors with the mean prevalence and lower and upper limits of 95% credible intervals with the values of the columns `mean`, `0.025quant` and `0.975quant` and the rows given by index. These vectors are included as columns in the `coop.df` data frame.
```{r}
coop.df <- as.data.frame(coop)
coop.df$prev_mean <- fit$summary.fitted.values[index, "mean"]
coop.df$prev_ll <- fit$summary.fitted.values[index, "0.025quant"]
coop.df$prev_ul <- fit$summary.fitted.values[index, "0.975quant"]
```

Now we create a map with the predicted prevalence (`prev_mean`) at the prediction locations `coop`.
```{r}
coop.sfdf <- st_as_sf(x = coop.df, 
                      coords = c("x", "y"),
                      crs = 4326)

tmap_mode(mode = "view")

tm_shape(coop.sfdf) +
  tm_dots(col = "prev_mean", size = .1)
```

Instead of showing the prevalence predictions at points, we can plot them using a raster. Here, the prediction locations `coop` are not on a regular grid; therefore, we need to create a raster with the predicted values using the `rasterize()` function. 

We transfer the predicted values `prev_mean` from the locations `coop.df` to the raster `ra` that we used to get the prediction locations with the following arguments.

`x = coop`: coordinates where we made the predictions,
`y = ra`: raster where we transfer the values,
`field = coop.df$prev_mean`: values to be transferred (prevalence predictions in locations coop),
`fun = mean`: to assign the mean of the values to cells that have more than one point.

```{r}
r_prev_mean <- rasterize(x = coop, 
                         y = ra, 
                         field = coop.df$prev_mean,
                         fun = mean)

tm_shape(r_prev_mean) +
  tm_raster(palette = "YlOrRd", alpha = .5)
```

We can follow the same approach to create maps with the lower and upper limits of the prevalence predictions.

#### Mapping exceedance probabilities

We can also calculate the exceedance probabilities of malaria prevalence being greater than a given threshold value that is of interest for policymaking. For example, we can be interested in knowing what are the probabilities that malaria prevalence is greater than 20%. Let  $p_i$ be the malarie prevalence at location ${\bf x}_i$. Then the probability that the malaria prevalence $p_i$ exceeds some value $c$ can be written $P(p_i > c)$.

This probability is calculated by subtracting $P(p_i \le c)$ from 1. With {inla} $P(p_i \le c)$ is obtained from the `inla.pmarginal()` function as
```{r, eval=FALSE}
1 - inla.pmarginal(q = c, marginal = marg)
```
where `marg` is the marginal distribution of the predictions, and `c` is the threshold value.

In our example, we can calculate the probabilites that malaria prevalence exceeds 20% as follows. First, we obtain the posterior marginals of the predictions for each location. These marginals are in the list object `fit$marginals.fitted.values[index]` where `index` is the vector of the indices of the stack `stk.full` corresponding to the predictions. In the previous section, we obtained these indices by using the `inla.stack.index()` function and specifying `tag = "pred"`.
```{r}
index <- inla.stack.index(stack = stk.full, tag = "pred")$data
```

The first element of the list, `fit$marginals.fitted.values[index][[1]]`, contains the marginal distribution of the prevalence prediction corresponding to the first location. The probability that malaria prevalence exceeds 20% at this location is given by
```{r}
marg <- fit$marginals.fitted.values[index][[1]]
1 - inla.pmarginal(q = .20, marginal = marg)
```

To compute the exceedance probabilities for all the prediction locations, we can use the `sapply()` function with two arguments. The first argument denotes the marginal distributions of the predictions (`fit$marginals.fitted.values[index]`) and the second argument denotes the function to compute the exceedance probabilities (`1- inla.pmarginal()`). Then the `sapply()` function returns a vector of the same length as the list `fit$marginals.fitted.values[index]`, where each element is the result of applying the function `1 - inla.pmarginal()` to the corresponding element of the list of marginals.
```{r}
excprob <- sapply(fit$marginals.fitted.values[index],
                  FUN = function(marg){1-inla.pmarginal(q = .20, marginal = marg)})

head(excprob)
```

Finally we rasterize these values and make a map.
```{r}
r_excprob <- rasterize(x = coop, 
                       y = ra, 
                       field = excprob,
                       fun = mean)

tm_shape(r_excprob) +
  tm_raster(palette = "RdPu", alpha = .5)
```

The map shows the probability that malaria prevalence exceeds 20% in The Gambia. This map quantifies the uncertainty relating to the exceedance of the threshold value 20%, and highlights the locations most in need of targeted interventions. In this map, locations with probabilities close to 0 are locations where it is very unlikely that prevalence exceeds 20%, and locations with probabilities close to 1 correspond to locations where it is very likely that prevalence exceeds 20%. Locations with probabilities around .5 have the highest uncertainty and correspond to locations where malaria prevalence is with equal probability below or above 20%.