# Pixelate (prediction + uncertainty)

https://arxiv.org/pdf/2005.11993.pdf

In brief, the function `pixelate()` groups predictions into a specified number of large pixels; computes the average uncertainty within each large pixel; then, for each large pixel, depending on its average uncertainty, either averages the predictions across it or across smaller pixels nested within it. These averaged predictions are then plotted.

The resulting plot of averaged predictions is selectively pixelated, similar to a photo that is deliberately pixelated to disguise a person’s identity. Areas of high average uncertainty in the pixelated plot are unresolved, while areas with high average certainty are resolved, similar to information poor versus rich regions of a satellite map.

This can take a few minutes.
```{r}
devtools::install_github("artaylor85/pixelate", build_vignettes = TRUE, dependencies = TRUE)
```

The {pixelate} package centres around a single function `pixelate()`. To use it and visualise its output following our examples, simply load and attach the pixelate package then read its vignette for both quick and detailed examples. 

In addition (and if you did not build the vignette upon installation), please see the documentation of `pixelate()` (accessed by `?pixelate`).

Follow the vignette.
```{r}
library(pixelate)
vignette("pixelate")
```

```{r}
head(SubSaharanAfrica_Pf_incidence)
```
